---
title: "Getting started with clonecensorweighting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with clonecensorweighting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(clonecensorweighting)
```

## Why clone-censor-weighting?

Suppose we want to know whether surgery improves survival in lung cancer
patients. Comparing patients who had surgery against those who did not, using a
naive analysis, can be badly biased. One reason is *immortal time bias*: to
appear in the "surgery" group, a patient has to survive long enough to actually
receive surgery. That guaranteed survival time gets unfairly credited to the
surgery group.

Clone-censor-weighting (CCW) is a technique used in target trial emulation to
address this. The idea has three parts, which also give the method its name:

- **Clone**: each patient is copied into every treatment strategy under
  comparison, so that at the start of follow-up everyone is compatible with
  every strategy.
- **Censor**: a clone is censored at the moment the patient's observed data
  stops being consistent with the strategy that clone was assigned to.
- **Weight**: inverse-probability-of-censoring weights correct for the
  artificial censoring introduced in the previous step.

This vignette walks through the pipeline end to end, using a lung cancer
dataset that ships with the package.

## The example data

The package includes `lungcancer`, a dataset of 200 lung cancer patients.

```{r}
data(lungcancer)
head(lungcancer)
```

The columns we rely on are:

- `id`: patient identifier
- `surgery`: whether the patient received surgery (`1`) or not (`0`)
- `timetosurgery`: time from baseline to surgery (`NA` if never operated on)
- `death`: whether the patient died (`1`) or not (`0`) — the outcome event
- `fup_obs`: observed follow-up time

The remaining columns (`sex`, `charlson`, `perf`, `stage`, `emergency`, `age`,
`deprivation`) describe patient characteristics that can be used as covariates
when estimating censoring weights.

We compare two strategies, and give them names we will reuse throughout:

```{r}
arms <- c("Control", "Surgery")
```

The **grace period** is central to this analysis. It is the window during which
a patient is still considered compatible with the "surgery" strategy. A patient
who has surgery within the grace period is consistent with the treated strategy;
one who does not is consistent with the control strategy. Here we use a grace
period of 90 days.

```{r}
grace_period <- 90
```

## Step 1: Clone the data across arms

`clone_arms()` copies the full dataset once per strategy and returns a list with
one data frame per arm.

```{r}
clones <- clone_arms(lungcancer, arms)

names(clones)
```

At this point the two data frames are identical copies of the original data. The
strategy-specific logic is applied in the next steps.

## Step 2: Build the policy and censoring logic

Two helper functions describe how each arm's emulated outcome, follow-up, and
censoring should be derived. They do not touch the data yet; they return the
rules (as expressions) that will be applied in Step 3.

`create_policy_A()` produces the emulated outcome and follow-up under the grace
period policy.

```{r}
policy <- create_policy_A(
  arms              = arms,
  treatment         = "surgery",
  time_to_treatment = "timetosurgery",
  grace_period      = grace_period,
  outcome           = "death",
  followup          = "fup_obs"
)
```

`create_censoring_logics_A()` produces the censoring indicator and the time at
which that indicator can first be determined.

```{r}
censoring <- create_censoring_logics_A(
  arms              = arms,
  treatment         = "surgery",
  time_to_treatment = "timetosurgery",
  grace_period      = grace_period,
  followup          = "fup_obs"
)
```

Each of these returns a nested list, one entry per arm. The policy rules create
the variables `.outcome` and `.fup`; the censoring rules create `.censoring` and
`.fup_uncensored`.

## Step 3: Apply the logic to the clones

`apply_logics()` evaluates the rules against each arm's data. We combine the
policy and censoring rules for each arm into a single set of new variables.

```{r}
logics <- list(
  Control = c(policy$Control, censoring$Control),
  Surgery = c(policy$Surgery, censoring$Surgery)
)

emulated <- apply_logics(clones, logics)

lapply(emulated, names)
```

Each arm's data frame now carries the emulated variables (`.outcome`, `.fup`,
`.censoring`, `.fup_uncensored`) alongside the original columns.

## Step 4: Assemble the final long-form data

`create_final_data()` turns each patient into a sequence of time intervals
(a counting-process, or "long", format). Internally it finds every event time,
splits each patient's follow-up at those times, and combines the outcome and
censoring information into one table per arm.

```{r}
final <- create_final_data(
  clones          = emulated,
  clone_followup  = ".fup",
  clone_outcome   = ".outcome",
  clone_censoring = ".censoring",
  col_ids         = "id"
)

head(final$Surgery)
```

Each row is now one patient-interval, bounded by `Tstart` and `Tstop`. Because a
single patient contributes several intervals, each arm has many more rows than
the original 200 patients:

```{r}
lapply(final, dim)
```

This long-form data is the input for the final steps of a CCW analysis:
estimating inverse-probability-of-censoring weights over time and fitting a
weighted survival model to compare the two strategies.

## Reading your own data

The example above uses the bundled `lungcancer` data. To run the same workflow
on your own data, it needs, at a minimum:

- an **identifier** column for each patient
- a binary **treatment** column, coded `0` / `1`
- a **time-to-treatment** column (numeric, `NA` for the untreated)
- a binary **outcome** column, coded `0` / `1`
- a numeric **follow-up time** column

The column *names* are up to you; you pass them to the arguments of the policy
and censoring helpers, as we did above. Any additional columns are carried along
and can serve as covariates.

`read_trial_data()` is a small convenience wrapper for reading such data from a
CSV file into a tibble. It does not impose any particular column structure; it
simply reads the file:

```{r, eval = FALSE}
my_data <- read_trial_data("path/to/your-data.csv")
```

## Summary

In this vignette we:

- motivated clone-censor-weighting with a lung cancer surgery question and the
  problem of immortal time bias,
- cloned the data across a control and a surgery arm with `clone_arms()`,
- described the grace-period policy and censoring rules with `create_policy_A()`
  and `create_censoring_logics_A()`,
- applied those rules with `apply_logics()`,
- and assembled a long-form analysis dataset with `create_final_data()`.

The remaining steps of a full CCW analysis — estimating censoring weights and
fitting a weighted outcome model — build directly on this long-form dataset.